RL-Musician: A Tool for Music Composition with Deep Reinforcement Learning

### 2019-10-03 docs/paper/paper.tex
docs: Write about related work

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating}.

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent can be either playing a note at the current time step or moving to the next time step. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method.

@@

Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested. The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or other methods (genetic algorithms, ant colony algorithms, evolutionary strategies \cite{salimans2017evolution}).

@@

%Reinforcement learning can be used for

Since algorithmic composition can be framed as training of a generative model, it sounds natural to try one of the most salient examples of generative models -- generative adversarial networks (GAN) \cite{goodfellow2014gan}. However, widespread notations for music assume sequences of discrete values but classical GAN work well only with continuous data, because gradient of discrete-valued functions is uninformative. Techniques originating from RL can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}. Namely, generator is trained with policy gradient method \cite{williams1992simple}. Such methodology is applied to various tasks including music composition \cite{limaguimaraes2017objective}.

@@

@@

\section{Further Improvements}
\label{sec:improvements}
% better evaluational rules (tonality);
% randomized initial observations;
% LSTM as actor model;
% evolutionary strategies instead of cross-entropy;
% start from supervisedly pre-trained weigths;

>>>>

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that can not be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}.

@@

Brief outline of the approach is as follows. There is an environment with 2D table representing piano roll which means that rows of the table correspond to consecutive notes and columns of the table correspond to time steps. An agent observes current time step and some previous time steps. An action of the agent can be either playing a note at the current time step or moving to the next time step. When the last time step of the piano roll is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model, i.e., a model that maps observation to probabilities of actions. These parameters are trained with cross-entropy method (alternatively, they can be trained with genetic algorithms, ant colony algorithms, or evolutionary strategies \cite{salimans2017evolution}).

@@

Further, sequential generation of composition by neural network implies that any well-studied architectures from sequence generation tasks can be plugged in and tested. The only difference is that now these neural networks can not be trained with variations of gradient descent, but can be trained with cross-entropy method or alternative methods.

@@

Reinforcement learning can be used for altering weights of recurrent neural networks trained to generate music sequentially \cite{jaques2016generating, kotecha2018bach}. The goal is to make generated pieces more structured and conformed with music theory rules. To define an environment, let its state be composed of recurrent neural networks states and previously played notes, let an action be an output for current time step and let reward depend on both evaluational rules and probability of output according to initial RNN. Rewards are granted immediately after a step and so DQN (Deep Q-Networks \cite{mnih2013playing}) are preferred over cross-entropy method as a training algorithm.

Usually, softmax activation function is used in the last layer of generative RNN. However, it is possible to consider a family of activation functions parametrized by one parameter $t$ such that output distribution can vary from atomic distribution concentrated at the most probable action ($t = 0$) to distribution returned with softmax activation ($t = 1$) and to uniform distribution ($t \to +\infty$). Also, input vector can be extended by introducing an additional part indicating origination of initial input (so called plan). Then an environment can be defined so that state is a pair of $t$ and plan, actions are changes in either $t$ or plan, and reward depends on evaluation of produced with these settings piece \cite{kumar2019polyphonic}.

Since algorithmic composition can be framed as training of a generative model, it sounds natural to try one of the most salient examples of generative models -- generative adversarial networks (GAN) \cite{goodfellow2014gan}. However, widespread notations for music assume sequences of discrete values but classical GAN work well only with continuous data, because gradient of discrete-valued functions is uninformative. Techniques originating from RL can be used to overcome this obstacle \cite{yu2016seqgan,hjelm2017boundary}. Namely, generator is trained with policy gradient method \cite{williams1992simple}. Such methodology is applied to various tasks and music composition is amongst them \cite{limaguimaraes2017objective}.

Nevertheless, not all methods for automated music generation use deep learning. Musical pieces itself can be generated with genetic algorithms \cite{felice2002genorchestra} where selection is based on scores returned by hand-written evaluation rules.

@@

Initially, cross-entropy method was developed for estimation of rare events probability \cite{rubinstein1997optimization}. However, it was found that it is also appropriate for solving optimization problems. More detailed discussion of cross-entropy method can be found in \cite{boer2005tutorial}.

@@

\section{Further Improvements}
\label{sec:improvements}
% better evaluational rules (tonality, rhythm);
% randomized initial observations;
% LSTM as actor model;
% evolutionary strategies instead of cross-entropy;
% start from supervisedly pre-trained weigths;
